---
title: "Lecture 02 Demo"
subtitle: "**PSTAT 100:** Summer Session A, 2025"
author: "**Instructor:** Ethan P. Marzban"
format: 
  pdf:
    header-includes:
      - \usepackage[margin = 1in]{geometry}
---

First we load our necessary packages.
```{r}
#| message: False

library(tidyverse)
library(reshape2)
```

Now, we read in the data. Remember that you will need to set the working directory first!
```{r}
enrollments <- read.csv("data/enrollments.csv")
```

## First Task: Extract Lower-Div Courses

We begin by extracting only the lower-division courses. There are a couple of different ways to do this; I'll do so using the following two-step procedure:

1)    Create a vector that stores the names of the lower-div courses
2)    Filter the `enrollments` dataframe to return only entries whose `Course` value is one of the lower-div courses

```{r}
## Step 1: create a vector to store the names of the lower-div courses
lwr_div <- c("PSTAT 5A", "PSTAT 5LS", "PSTAT 5H", "PSTAT 8", "PSTAT 10")

## Step 2: Filter
lwr_div_courses <- enrollments %>% filter(Course %in% lwr_div)
```


## Second Task: Melt

Now, we need to melt our `lwr_div_courses` dataframe. As a reminder:

-   There are two colvars: `Course` and `Title`
-   We want the remaining column headers to become values of a single `Quarter` variable
-   We want the values to concactenate into a single `Enrollments` variable

Using the `melt()` function from the `reshape2` package:

```{r}
lwr_div_molten <- lwr_div_courses %>%
  melt(
    id.vars = c("Course", "Title"),
    variable.name = "Quarter",
    value.name = "Enrollment"
  )

lwr_div_molten %>% head()
```

## Third Task: Generate a Plot

```{r}
#| warning: False

lwr_div_molten %>% ggplot(aes(x = Quarter, y = Enrollment)) + 
  geom_point(aes(colour = Course)) +
  geom_line(aes(colour = Course, group = Course)) +
  theme_minimal() +
  scale_x_discrete(limits = rev) +
  ggtitle("Lower Div Enrollments over Time") 
```