---
title: "Lab 10: Lab-ubu"
subtitle: "PSTAT 100, Summer Session A 2025 with Ethan P. Marzban"
author:
  - MEMBER 1 (NetID 1)
  - MEMBER 2 (NetID 2)
  - MEMBER 3 (NetID 3)
date: "July 24, 2025"
date-format: long
format: 
  pdf:
    mathspec: true
    header-includes:
      - \usepackage[margin = 0.9in, 
                    top = 0.8in, 
                    bottom = 1.1in]{geometry}
      - \usepackage[headsepline]{scrlayer-scrpage}
      - \lohead{PSTAT 100, Summer Session A 2025}
      - \cohead{ \ Lab10}
      - \rohead{\copyright Ethan P. Marzban}
      - \newcommand{\mat}[1]{\mathbf{#1}}
      - \newcommand{\tmat}[1]{\mathbf{#1}^{\mathsf{T}}}
      - \newcommand{\vect}[1]{\vec{\boldsymbol{#1}}}
      - \newcommand{\tvect}[1]{\vec{\boldsymbol{#1}}^{\masthf{T}}}
---

```{r setup, echo = F}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = F}
## optional code chunk;
## gives shortcut for boldface colored text,
## able to be rendered in both PDF and HTML

bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```


# Required Packages

```{r, message = F, warning = F}
library(ottr)        # for checking test cases (i.e. autograding)
library(pander)      # for nicer-looking formatting of dataframe outputs
library(tidyverse)   # for graphs, data wrangling, etc.
```


# Logistical Details

:::{.callout-note}
## **Logistical Details**

-   This lab is due by **11:59pm on Wednesday, July 30, 2025**.

-   Collaboration is allowed, and encouraged!
    -   If you work in groups, list ALL of your group members' names and NetIDs (not Perm Numbers) in the appropriate spaces in the YAML header above.
    -   Please delete any "MEMBER X" lines in the YAML header that are not needed.
    -   No more than 3 people in a group, please.
    
-   Ensure your Lab properly renders to a `.pdf`; non-`.pdf` submissions will not be graded and will receive a score of 0.

-   Ensure all test cases pass (test cases that have passed will display a message stating `"All tests passed!"`)
:::

# Lab Overview and Objectives

Welcome to another PSTAT 100 Lab! In this lab, we will cover the following:

-   Clustering in `R`
-   Missing Data


# Part I: Clustering

_**Labubus**_ are a moderately-sized plush doll that has completely taken the online market recently. If you'd like to read more about the history of Labubus, feel free to consult [This NPR Article](https://www.npr.org/2025/06/18/g-s1-72939/what-is-labubu-pop-mart-explained). 

For the purposes of this lab, we'll explore a simulated dataset pertaining to _Labubus_. Specifically, we are imagining that someone has collectd 61 labubus and recorded the following attributes:

-   The height of the _Labubu_ (in centimeters)
-   The weight of the _Labubu_ (in grams)
-   The `RBG` color specification of the _Labubu_ (split across three columns; one for `R`, one for `G`, and one for `B`)

Additionally, it is known that each _Labubu_ falls into one of six styles: Soymilk (`SM`), Lychee Berry (`LB`), Green Grape (`GG`), Sea Salt Coconut (`SSC`), Toffee (`T`), and Sesame Bean (`SB`) (names taken from [this](https://www.popmart.com/us/products/675/the-monsters-exciting-macaron-vinyl-face-blind-box) listing). 

The full _Labubu_ dataset can be found in the `labubu.csv` file, located in the `data/` subfolder.

```{r}
labubu <- read.csv("data/labubu.csv")
```

\

:::{.callout-important}
## **Question 1**

Create a data frame called `labubu_num` that contains only the numerical columns from the `labubu` dataset. \

`r bfcolor("Solution:", "blue")`

```{r}
## replace this line with your code
```

`r bfcolor("Answer Check:", "blue")`
```{r, message = F}
# DO NOT EDIT THIS LINE
invisible({check("tests/q1.R")})
```

:::

\

Let's see what happens if we try to visually classify based on only two variables. \

:::{.callout-important}
## **Question 2**

Generate a scatterplot of the Labubus' heights (on the vertical axis) against their weights (on the horizontal) axis. Ensure your plot is presentation-quality. How many clusters do you see on the plot? **Don't** rely on the `type` column from the `labubu` dataframe - only use the plot you just generated! \

`r bfcolor("Solution:", "blue")`

```{r}
## replace this line with your code
```

_Replace this line with your answer_ \

`r bfcolor("Answer Check:", "blue")`

There is no autograder for this question; your TA will manually check that your answers are correct.

:::

Hm, our number of clusters doesn't appear to agree with the number of different _Labubu_ types we know to be present in the dataset. This _could_ mean that two or more of the _Labubu_ types are very similar to one another - however, it's dangerous to make that conclusion based solely on our work in Question 2 since we ignored several variables from the dataset. Let's remedy that!

\


:::{.callout-important}
## **Question 3**

Perform PCA on the `labubu_num` dataframe, taking care to _standardize_ (not only mean-center) each column. Then, use this to generate a plot of the first two latent factors; how many clusters do you see on this PC plot? \

`r bfcolor("Solution:", "blue")`

```{r}
## replace this line with your code
```

_Replace this line with your answer_ \

`r bfcolor("Answer Check:", "blue")`

There is no autograder for this question; your TA will manually check that your answers are correct.

:::

\

Wow- the clusters on this graph are pretty distinct! Nevertheless, let's run a _K_-means clustering algorithm on our PC-projected dataset. As a reminder, in `R` we use the `kmeans()` function to run the _K_-means clustering algorithm; take a look at the help file for this function before moving onto the next question on this lab.  \

:::{.callout-important}
## **Question 4**

**Part (a)**

Pass the matrix of the first two PCs into a call to `kmeans()`, and assign the result to a variable called `kmeans_labubu`. Use 6 clusters. **Make sure to set your seed to 100 before beginning; otherwise the autograder may not work.** \

`r bfcolor("Solution:", "blue")`

```{r}
set.seed(100)
## replace this line with your code
```


`r bfcolor("Answer Check:", "blue")`

```{r, message = F}
# DO NOT EDIT THIS LINE
invisible({check("tests/q4a.R")})
```

\

**Part (b)**

Re-do the PC plot from Question 3, now coloring based on the clusters identified by `kmeans()`.


```{r}
## replace this line with your code
```

`r bfcolor("Answer Check:", "blue")`

There is no autograder for this question; your TA will manually check that your answers are correct.

:::

Assuming you correctly completed the questions above, the following code should produce a cross-tabulation of each _Labubu_'s classification (as determined by _K_-means) and its true type (as classified by the `type` variable from the original `labubu` dataframe):

```{r}
#| echo: True
#| error: True
#| message: False

labubu %>%
  mutate(class = kmeans_labubu$cluster) %>%
  group_by(type, class) %>%
  summarise(count = n()) %>%
  pivot_wider(names_from = type,
              values_from = count,
              values_fill = 0) %>%
  pander()
```


:::{.callout-important}
## **Question 5**

Has _K_-means classified along the different types, as encoded by the `type` variable in the original `labubu` dataframe? Are there any two _Labubu_ types that are so similar to each other that they are indistinguishable (by _K_-means) from one another? How can you tell?  \

`r bfcolor("Solution:", "blue")`

_Replace this line with your answer_ \


`r bfcolor("Answer Check:", "blue")`

There is no autograder for this question; your TA will manually check that your answers are correct.

:::


\

Something that is important to note is that there is a bit of randomness in the _K_-means algorithm. As such, let's see how this manifests itself in this particular dataset.


:::{.callout-important}
## **Question 6**

Re-run the _K_-means clustering algorithm, but this time **set your seed to 20 at the start**. Produce a plot like you did in Question 4(b), coloring points by the _new_ classifications. Comment on any differences between this plot and the plot from 4(b) above. \


```{r}
set.seed(20)
## replace this line with your code
```

_Replace this line with your answer_ \

`r bfcolor("Answer Check:", "blue")`

There is no autograder for this question; your TA will manually check that your answers are correct.

:::

\

So, there is some variability in the product of the _K_-means clustering algorithm. Again, recall that the _K_-means algorithm is only optimal in a local sense - in some cases (like with this particular simulated _Labubu_ dataset), the final configurations may be dependent on the seed you set initially. 


# Submission Details

Congrats on finishing this PSTAT 100 lab! Please carry out the following steps:

:::{.callout-note}
## **Submission Details**

1)    Check that all of your tables, plots, and code outputs are rendering correctly in your final `.pdf`.

2)    Check that you passed all of the test cases (on questions that have autograders). You'll know that you passed all tests for a particular problem when you get the message "All tests passed!".

3)    Submit **ONLY** your `.pdf` to Gradescope. Make sure to **match ALL pages to the ONE question on Gradescope**; failure to do so will incur a penalty of 0.1 points.
:::


